---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document:
    keep_md: true
---



```{r, echo=FALSE}
setwd("d:/tmp/Dropbox/Edu/Coursea/DataScienceSpecialization/ReproducibleResearch/RepData_PeerAssessment1/")

#lets get course data
#DataUrl<-'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip'
#download.file(DataUrl,destfile = './data.zip')
unzip('./activity.zip')
#unlink('./activity.zip') #delete zip file
```


## Loading and preprocessing the data


```{r}
DataUrl<-file.path(getwd(),"activity.csv")

Data<-read.table(DataUrl,header = T, sep = ",")
names(Data)

#lets get some proper dates
Data$date<-as.POSIXct(paste(Data$date), format="%Y-%m-%d")

```


## What is mean total number of steps taken per day?

```{r}
NoOfDays<-length(unique(Data$date))
#aggregate by days
DailyWalk<-aggregate(steps ~ date,Data,sum)
summary(DailyWalk$steps,na.rm = T)

```

Data was taken over `r NoOfDays` days, with mean of `r mean(DailyWalk$steps,na.rm = T)` and median of `r median(DailyWalk$steps,na.rm = T)` steps taken per day. 

Data histogram suggest that activities vary from day to day

```{r}
library(ggplot2)


HistPlot<-ggplot(DailyWalk, aes(steps)) + geom_histogram(aes(y = ..density..), binwidth = 500) + labs(title = "Steps per day")
HistPlot

```



## What is the average daily activity pattern?

Lets compare average number of steps, across all days, in each interval. 

```{r}
#aggregate by days
DailyWalk<-aggregate(steps ~ interval,Data,mean)
summary(DailyWalk$steps,na.rm = T)

```


Represent it as a time series.

```{r}
ggplot(DailyWalk, aes(interval, steps))+geom_line()

TimeofMaxActivity<-DailyWalk$interval[DailyWalk$steps==max(DailyWalk$steps)]
```

Across all days peak of activity (no of steps), is around `r TimeofMaxActivity` interval.


## Imputing missing values


```{r}

#find N/As
MissingValues<-sum(!complete.cases(Data))
```

We got a total of `r MissingValues` ()missing records. 

We will replace NAs with average value for particular interval.

```{r}
library(plyr) #arrange
MisingValueMask<-!complete.cases(Data)

NAsFiller<-merge(Data[MisingValueMask,c("date","interval")],DailyWalk,by="interval",sort=F )
NAsFiller<-arrange(NAsFiller,date)

DataCorrected<-Data
#create a set without NAs
DataCorrected[MisingValueMask,c("steps")]<-NAsFiller[,c("steps")]
```

Lets compare how repacement of NAs changes mean and median total number of steps taken per day.

```{r}
#aggregate by days
DailyWalk<-aggregate(steps ~ date,DataCorrected,sum)
summary(DailyWalk$steps,na.rm = T)

#replog ggplot with upgraded dataframe
HistPlot %+% DailyWalk
```

**DataCorrected** mean is `r mean(DailyWalk$steps,na.rm = T)` and median is `r median(DailyWalk$steps,na.rm = T)` steps taken per day. 
Filling in NAs has removed small skew in the data, with mean and median bein now almost equal


## Are there differences in activity patterns between weekdays and weekends?

Lets identify weeekdays and weekends first.

```{r}
DayOfWeek<-weekdays(Data$date, abbreviate=T)

#define weekend
Data$DayType<-"weekday"
Data$DayType[DayOfWeek %in% c("Sat","Sun")]<-"weekend"
```

Lets plot and compare.

```{r}
ggplot(aes(x=interval,y=steps),data = Data)+geom_line(aes(colour = DayType))+  theme_bw()+facet_wrap(~DayType,nrow=2) +  labs(title = "Activity comparison between weekdays and weekends",legend.position = "none")  

```

There is difference in activities between weekends and weekdays.

